---
title: Spatially explicit models for distance sampling data
author: David L Miller
subtitle: "Practical 5: Prediction and variance"
date: 8-12 June 2020

---

## Aims

By the end of this practical, you should feel comfortable:

- Making a prediction using the `predict()` function
- Summing the prediction cells to obtain a total abundance for a given area
- Plotting a map of predictions
- Knowing when to use `dsm.var.prop` and when to use `dsm.var.gam`
- Estimating variance for a given prediction area
- Estimating variance per-cell for a prediction grid
- Interpreting the `summary()` output for uncertainty estimates
- Making maps of the coefficient of variation in R


## Loading the packages and data

```{r load-packages}
library(knitr)
library(dsm)
library(ggplot2)
```

Load our models and data:

```{r load-models}
load("spermwhale.RData")
# load models
load("dsms.RData")
load("dsms-checked.RData")
# rename one model
dsm_nb_xy_ms <- dsm_nb_xy_ms_p1_4
```

## Prediction data

Our prediction data lives in the `predgrid` object. This data was prepared from various sources and the since some of the covariates were dynamic (SST, NPP and EKE), the dates for these were arbitrarily chosen as 2 June 2004, or Julian date 153.

If you're preparing this data yourself, the [`rerddap` package](https://cran.r-project.org/web/packages/rerddap/) can be used to obtain marine data from NOAA for both segments and predictions.

We can take a look at our data using `head()` to see the first few rows:

```{r inspect-preddata}
head(predgrid)
```

As we saw in the slides, the `predict()` function will fill in the variables in the model formula with these values to obtain the predictions at each cell.

For example using the model `dsm_nb_xy_ms`:

```{r makepred1}
pp <- predict(dsm_nb_xy_ms, predgrid, off.set=predgrid$off.set)
```

This is just a list of numbers -- the predicted abundance per cell. We can sum these to get the estimated abundance for the study area:

```{r predsum}
sum(pp, na.rm=TRUE)
```

We can also plot this to get a spatial representation of the predictions:
```{r predsp, fig.cap="Predicted surface for abundance estimates with bivariate spatial smooth along with environmental covariates."}
# assign the predictions to the prediction grid data.frame
predgrid$Nhat_nb_xy <- pp

# plot!
p <- ggplot(predgrid) +
  geom_tile(aes(x=x, y=y, fill=Nhat_nb_xy, width=10*1000, height=10*1000)) +
  coord_equal() + 
  # better colour scheme  
  scale_fill_viridis_c() +
  # cleaner plot
  theme_minimal() +
  # set the label for the fill colour
  labs(fill="Abundance")
print(p)
```

Use the chunks above as a template and make predictions for the other models you saved in the previous exercises. In particular, compare the models with only spatial terms to those with environmental covariates included.

## Variance estimation

Depending on the covariates in the model, we can use either `dsm.var.prop` or `dsm.var.gam`. Remember:
- `dsm.var.gam`
  - assumes spatial model and detection function are independent
- `dsm.var.prop`
  - propagates uncertainty from detection function to spatial model
  - only works for `count` models
  - covariates can only vary at segment level

If there are no covariates in the detection function, use `dsm.var.gam` since there's no covariance between the detection function and the spatial model.
  
`dsm_nb_xy_ms` doesn't include any covariates at the observer level in the detection function, so we can use `dsm.var.gam` to estimate the uncertainty.

```{r varest}
# now estimate variance
var_nb_xy_ms <- dsm.var.gam(dsm_nb_xy_ms, predgrid, 
                             off.set=predgrid$off.set)
```
To summarise the results of this variance estimate:
```{r varest-summary}
summary(var_nb_xy_ms)
```

Try this out for some of the other models you've saved.

## Plotting

We can plot a map of the coefficient of variation, but we first need to estimate the variance per prediction cell, rather than over the whole area. 

```{r per-cell-var}
# use the split function to make each row of the predictiond data.frame into
# an element of a list
predgrid_var_split <- split(predgrid, 1:nrow(predgrid))
var_split_nb_xy_ms <- dsm.var.gam(dsm_nb_xy_ms, predgrid_var_split, 
                                   off.set=predgrid$off.set)
```
Now we have the per-cell coefficients of variation, we assign that to a column of the prediction grid data and plot it as usual:

```{r varest-map-obs, fig.cap="Uncertainty (CV) in prediction surface from bivariate spatial smooth with environmental covariates.  Sightings overlaid."}
cv <- sqrt(var_split_nb_xy_ms$pred.var)/unlist(var_split_nb_xy_ms$pred)
predgrid$CV <- cv
p <- ggplot(predgrid) +
  geom_tile(aes(x=x, y=y, fill=CV, width=10*1000, height=10*1000)) +
  scale_fill_viridis_c() +
  coord_equal() +
  geom_point(aes(x=x,y=y, size=count), alpha=0.4,
             data=dsm_nb_xy_ms$data[dsm_nb_xy_ms$data$count>0,]) +
  theme_minimal()
print(p)
```

Note that here we overplot the segments where sperm whales were observed (and scale the size of the point according to the number observed), using `geom_point()`. The `alpha=` argument makes the points semi-transparent.

Try this with the other models you fitted and see what the differences are between the maps of coefficient of variation.


## Extra credit

- Try plotting the effort over a CV plot, using the data in the model.
- The functions `dsm.var.prop` and `dsm.var.gam` can accept arbitrary splits in the data, not just whole areas or cells. Make a `list` with two elements: one a `data.frame` of all the cells with $y>0$ and one with $y\leq 0$. Estimate the variance for these regions. Note that you'll need to sum the offsets for each area to get the correct value to supply to `off.set=...`.

